SmartCheck: Automatic and Efficient Counterexample Reduction and Generalization

### 2014-03-20 paper/paper.tex
paper wibbles.

@@

\begin{abstract}
QuickCheck, developed by Claessen and Hughes, is a powerful library for automatic test-case generation.  Because QuickCheck performs random testing, some of the counterexamples discovered are very large.  QuickCheck provides an interface for the user to write shrink functions to attempt to reduce the size of counterexamples discovered.  Hand-written implementations of shrink can be complex, inefficient, and consist of significant boilerplate code.  Furthermore, shrinking is only one aspect in debugging: counterexample generalization is the process of extrapolating from individual counterexamples to a class of counterexamples, often requiring a flash of insight from the programmer.  To improve counterexample reduction and generalization, we introduce \emph{SmartCheck}.  SmartCheck is a debugging tool that reduces algebraic data using generic search heuristics to efficiently find smaller counterexamples independent of the program or property being tested.  In addition to shrinking, SmartCheck also automatically generalizes counterexamples to formulas representing them.  SmartCheck has been implemented for Haskell and is freely available.
\end{abstract}

@@

In addition, shrinking is one side of the coin when it comes to making
counterexamples more understandable: the other side is extrapolation from
individual counterexamples to a class of counterexamples.  This leap of
abstraction is often implicitly made by the programmer in determining the reason
why counterexamples fail the property.  For example, the following is a
relatively small counterexample returned when QuickChecking a property in (a
bug-injected version of) Xmonad, a popular X11 window manager written in
Haskell~\cite{xmonad}:
%
\medskip%
\begin{sloppypar}
%\footnotesize
\noindent%
\ttp{StackSet \{current = Screen \{workspace = Workspace \{tag = NonNegative
      \{getNonNegative = 0\}, layout = -1, stack = Just (Stack \{focus = 'S', up =
        "", down = ""\})\}, screen = 1, screenDetail = 1\}, visible = [Screen
    \{workspace = Workspace \{tag = NonNegative \{getNonNegative = 2\}, layout = -1,
        stack = Nothing\}, screen = 2, screenDetail = -1\},Screen \{workspace =
      Workspace \{tag = NonNegative \{getNonNegative = 3\}, layout = -1, stack =
        Nothing\}, screen = 0, screenDetail = -1\}], hidden = [Workspace \{tag =
      NonNegative \{getNonNegative = 1\}, layout = -1, stack = Just (Stack \{focus
        = `\char`\\NUL', up = "", down = ""\})\},Workspace \{tag = NonNegative
      \{getNonNegative = 4\}, layout = -1, stack = Just (Stack \{focus = 'I', up =
        "", down = ""\})\}], floating = fromList []\} }
\end{sloppypar}
\medskip%
\noindent
(This counterexample uses Haskell's default \ttp{Show} instances, which uses
record syntax.)  Programmers should be familiar with having to debug
a ``wall of text'' as shown above.  What if instead a
formula like the following were returned, stating that for any well-typed
values $x_0$, $x_1$, $x_2$, and $x_3$, tested, a counterexample is found?

@@

As shown in the motivating example in Section~\ref{sec:example}, SmartCheck
can shrink more than an order-of-magnitude faster than efficient hand-written
shrink implementations for some properties and programs.  As well, the time
required to shrink scales linearly with the size of the counterexample---other
common shrink implementations often scale in polynomial time or worse.
(Shrinking time is important in real-world uses; see
Section~\ref{sec:experiments} for examples.)

@@

A more clever programmer that understands the semantics of QuickCheck's
\ttp{shrink} implementation,\footnote{The following approach was suggested to
  the author by John Hughes.} defines a \ttp{shrink} instance as follows:
%
\begin{code}
shrink (T w i0 i1 i2 i3) = map go xs
  where xs = shrink (w, i0, i1, i2, i3)
        go (w', i0', i1', i2', i3')
           = T w' i0' i1' i2' i3'
\end{code}
%
\noindent
This ``tuple shrink'' definition does not suffer the same shortcomings: it
shrinks a value even if it contains an empty list, and the combinatorial blowup
of shrink candidates is avoided, since a pair \ttp{(a,b)} is shrunk by
attempting to shrink \ttp{a} while holding \ttp{b} constant, then attempting to
shrink \ttp{b} while holding \ttp{a} constant (and is generalized to larger
tuples).  Thus, each argument to \ttp{T} is independently shrunk.

@@

@@

To motivate the benefits of SmartCheck reduction algorithm, consider
Table~\ref{table:results} and the corresponding graphs in
Figure~\ref{fig:graphs}, giving the results for using QuickCheck and SmartCheck
on the program shown in Figure~\ref{fig:initial} (we omit SmallCheck, since as
noted, it is not even feasible for the example with default instances).  We
graph the final counterexample size and execution time.  Out of 2000 runs, we
show the number of runs resulting in a shrunk counterexample of a given size (by
counting \ttp{N} constructors), and we show the number of runs executing within
a given time limit (in seconds).  Note that we provide two execution-time
graphs, since time scales differ dramatically (we use a logrithmic scale for the
$y$ axis in one graph and a linear scale in the other).  In the table, we show
the mean, median, median, and the results at the 95th percentile values.  (The
distributions produced by these values are not necessarily Gaussian, so we omit
the standard deviation.)

@@

In regards to generalization, after shrinking the counterexample, SmartCheck
will also generalize it.  Note that for any \ttp{Word8} value in the first
argument of \ttp{T}, the counterexample still holds.  For example, SmartCheck
might return a counterexample of the form
%
\begin{code}
forall x . T x [-20000] [-20000] [] []
\end{code}

Finally, these performance results for SmartCheck are fairly generalizable to
other types and properties, since the algorithm is not dependent on the specific
program or property being checked.  However, the results are particularly
pronounced for counterexamples that are particularly ``deep'' or ``broad'' in
terms of their tree structure.  For programs and properties for which small
counterexamples are easily found, QuickCheck's shrinking is perfectly
acceptable.  Furthermore, user-defined shrink methods may vary significantly
based on the property and counterexample type, so it is not always clear which
implementation to compare against.

@@

class Arbitrary a => SubTypes a where
  size    :: a -> Size
  index   :: a -> Idx -> Maybe SubVal
  replace :: a -> Idx -> SubVal -> a
  constr  :: a -> String
  constrs :: a -> [String]
  opaque  :: a -> Bool
  subVals :: a -> Tree SubVal
\end{code}
%
\noindent
The \ttp{SubTypes} type class requires QuickCheck's \ttp{Arbitrary} as a
super-class.  \ttp{SubTypes} has the following methods:
\begin{itemize}
\item \ttp{size} returns the size of a value---intuitively, the number of
  constructors contained within it,
\item \ttp{index} returns a \emph{sub-value} at a breadth-first index in a
  value,\footnote{Rather than indexes, an alternative representation is to use a
    zipper data structure~\cite{zipper} to traverse data.  We have chosen
    explicit indexes to write simple tail-recursive algorithms that can easily
    be transcribed to imperative languages.}
\item \ttp{replace} replaces a sub-value at a particular focus (returning the
  original value if the index is out-of-bounds),
\item \ttp{constr} returns a string representation of the constructor tagging
  the value,
\item \ttp{constrs} returns the list of all possible constructors from the
  value's type,
\item \ttp{opaque} is false when the type of the value is an ``interesting
  type''; informally, this is a type other than a primitive type like \ttp{Int}
  or \ttp{Char}.  See Section~\ref{sec:base} for a full discussion.
\item \ttp{subVals} returns a tree of all non opaque-type sub-values.  A tree is a
  value with the type
%
  \begin{code}
data Tree a = Node \{ rootLabel :: a
                   , subForest :: [Tree a] \}
  \end{code}
%
\end{itemize}

@@

subVals (B (B L\sub{0} L\sub{1}) L\sub{2}) =
  Node (B L\sub{0} L\sub{1}) [Node L\sub{0} [], Node L\sub{1} []]
       (Node L\sub{1} [])
\end{code}
%
\noindent
The \ttp{SubVal} type is an existential data type, used as a generic container
for sub-values from a counterexample.  We will sometimes refer to the unwrapped
value returned by \ttp{index a i} as the \emph{$i$th sub-value of $a$}, so for
example, (\ttp{B\sub{4} L\sub{6} L\sub{8}}) is the 4th sub-value of \ttp{tree}.
An invariant of \ttp{index} is that for any value \ttp{a}, and for the smallest
$i \geq 0$ such that
%
\begin{code}
index a i == Nothing
\end{code}
%
\noindent
then for all $0 \leq j < i$,
%
\begin{code}
index a j /= Nothing
\end{code}
\noindent
We use this invariant as a termination case in recursive algorithms over the
sub-values of a value.

@@

\section{Shrinking Data}\label{sec:shrinking}
In this section, we describe how to efficiently and generically shrink algebraic
data values.  Before we begin, recall the basic idea behind the \ttp{shrink}
method of the \ttp{Arbitrary} class: generate a list of values, each of which is
smaller than the current counterexample.  Each of the new values generated
may not bear any relationship to the original counterexample other than being
smaller.  SmartCheck pursues an approach that searches for smaller but
structurally similar counterexamples, as we make precise below.

@@

newVals :: Size -> Int -> SubVal -> IO [SubVal]
newVals sz tries (SubVal a) =
  replicateM tries s where
  s = liftM SubVal (sizedArbitrary sz a)

@@

test :: SubTypes a => a -> Idx -> [SubVal]
     -> (a -> Property) -> Maybe a
test cex idx vs prop = go vs
  where
  go []      = Nothing
  go (v:vs') =
    let cex' = replace cex idx v in
    if pass prop cex' then go vs'
      else Just cex'
  \end{code}
  \caption{Counterexample reduction algorithm.\label{fig:reduction}}
\end{figure}

Figure~\ref{fig:reduction} shows the reduction algorithm.  The function
\ttp{reduce} takes flags to customize the algorithm's behavior, a counterexample
\ttp{cex}, and the property \ttp{prop}.  The reduction begins at the first
proper sub-value of \ttp{cex}; call it \ttp{v}.  When the index \ttp{idx}
becomes out-of-bounds and returns \ttp{Nothing}, the algorithm terminates.
Otherwise, a list of new random values are generated.
%
\begin{code}
sizedArbitrary :: SubTypes a => Size -> a -> IO a
\end{code}
%
\noindent
generates a new value \ttp{v'} having the same type as \ttp{v} and that is
strictly smaller (with respect to the \ttp{size} method) than \ttp{v}.  Just
like QuickCheck's \ttp{arbitrary} method, \ttp{sizedArbitrary} generates
successively larger counterexamples when generating new values with which to
replace a sub-value.

@@

\subsubsection{Variant Counterexample Hypothesis}
A motivation for the design of the reduction algorithm is something we call the
\emph{variant counterexample hypothesis}: in the search space of possible values
from a given type \ttp{T}, if a known counterexample \ttp{cex} is a variant
\ttp{v} of \ttp{T}, then it is most probable that other counterexamples are also
from variant \ttp{v}.  Recall that the algorithm begins at the \emph{first}
sub-value of the counterexample rather than the 0th sub-value.  The 0th
sub-value of \ttp{cex} is \ttp{cex} itself.  No invariant of the algorithm would
be violated by beginning with the 0th sub-value, and in particular, the
algorithm would still terminate.  However, we begin with a strict sub-value
because of the hypothesis.  As an example supporting the hypothesis, consider a
property about unintended variable capture over a language's parse tree
represented by a sum type with constructors for module imports, function
definitions, and global-variable assignments, respectively.  A function
definition counterexample can only be reduced to smaller function definition
counterexamples, the only construct in which variable capture is possible.

@@

@@

  testHole (SubVal a) = do
    a' <- cast a :: Maybe a
    if pass prop a' then Nothing else Just a'
\end{code}
  \caption{Reduction algorithm with the sub-value counterexample optimization.}
  \label{fig:reduce0}
\end{figure}

@@

\subsection{Universal Sub-Value Generalization}\label{sec:universal}
Recall from the end of Section~\ref{sec:example} the quantified formula
%
\begin{code}
forall x . T x [-20000] [-20000] [] []
\end{code}
%
\noindent
characterizing a set of counterexamples.  Another example using the calculator
language from Section~\ref{sec:subval} is to generalize a counterexample like
%
\begin{code}
Div (Add (C 7) (C 3)) (Add (C (-5)) (C 5))
\end{code}
%
\noindent
by
%
\begin{code}
forall x . Div x (Add (C (-5)) (C 5))
\end{code}
%
\noindent
since any dividend results in divide-by-zero for the given divisor.  Not only do
the generalizations assist the programmer's insight, but they reduce the sheer
size of the counterexample.  We call the kind of formula just shown
\emph{universal sub-value generalization} and it is implemented in SmartCheck.

@@

\begin{figure}
  \begin{code}
subTrees :: SubTypes a => a -> Idx -> [Idx] -> Bool
subTrees cex idx = any (subTree cex idx)

@@

allFail :: SubTypes a => ScArgs -> a -> Idx
  -> [SubVal] -> (a -> Property) -> Bool
allFail args cex idx vs prop =
  length res >= scMinForall args && and res
  where
  res  = mapMaybe go vs
  go   = fail prop . replace cex idx
  \end{code}
  \caption{Universal sub-value generation algorithm.}
  \label{fig:universal}
\end{figure}

The extrapolation algorithm is shown in Figure~\ref{fig:universal}.  The
algorithm is similar to the reduction algorithm in Figure~\ref{fig:reduction},
and in the implementation, the algorithms are generalized and combined.  The
function \ttp{extrapolate} returns a list of indexes to be generalized in the
original counterexample.  In the recursive function \ttp{extrapolate'}, there is
a function guard with a call
%
\begin{code}
subTree cex idx0 idx1
\end{code}
%
\noindent
where \ttp{subTree} has the type
%
\begin{code}
subTree :: SubTypes a => a -> Idx -> Idx -> Bool
\end{code}
%
\noindent
is true if in \ttp{cex}, the value at index \ttp{idx0} is a child of index
\ttp{idx1} in a tree representation of \ttp{cex} (i.e., \ttp{subVals cex}).  The
\ttp{subTrees} guard prevents the algorithm from trying to generalize sub-values
that are abstracted away already since their parents have been generalized.

New sub-values are generated by \ttp{newVals}, shown in
Figure~\ref{fig:reduction}.  Like in the reduction algorithm, we bound the size
of values to generate by the flag \ttp{scMaxSize}, which is independent of the
size of the particular sub-value.  The flag \ttp{scMaxForall} is the analogue of
the \ttp{scMaxReduce} flag, determining the number of values generated in trying
to generalize a value.

The function \ttp{allFail} takes a counterexample \ttp{cex}, an index into
\ttp{cex}, a list of new sub-values, and a property.  It returns true if no new
values satisfy the property.  The function
%
\begin{code}
fail :: (a -> Property) -> a -> Maybe Bool
\end{code}
%
\noindent
is roughly the dual of \ttp{pass} in the reduction algorithm: (\ttp{fail
  prop cex}) returns (\ttp{Just True}) if \ttp{cex} passes \ttp{prop}'s
precondition but fails the property; (\ttp{Just False}) if \ttp{cex}
non-trivially satisfies \ttp{prop}, and \ttp{Nothing} if \ttp{cex} fails
\ttp{prop}'s precondition.  The flag \ttp{scMinForall} is the minimum number
of \ttp{Just False} results required from \ttp{fail} to extrapolate from failed
tests to a universal claim.

@@

Regarding our implementation, we have not profiled the performance of our
implementation of SmartCheck, so additional performance gains might be possible.
In particular, we perform significant computations over tree-structured data,
and a generic zipper~\cite{zipper}, which we do not currently implement, could
improve performance.

>>>>

\newcommand{\undef}[1]{{\bf #1}}

@@

\begin{abstract}
QuickCheck, developed by Claessen and Hughes, is a powerful library for
automatic test-case generation.  Because QuickCheck performs random testing,
some of the counterexamples discovered are very large.  QuickCheck provides an
interface for the user to write shrink functions to attempt to reduce the size
of counterexamples.  Hand-written implementations of shrink can be complex,
inefficient, and consist of significant boilerplate code.  Furthermore,
shrinking is only one aspect in debugging: counterexample generalization is the
process of extrapolating from individual counterexamples to a class of
counterexamples, often requiring a flash of insight from the programmer.  To
improve counterexample reduction and generalization, we introduce
\emph{SmartCheck}.  SmartCheck is a debugging tool that reduces algebraic data
using generic search heuristics to efficiently find smaller counterexamples.  In
addition to shrinking, SmartCheck also automatically generalizes counterexamples
to formulas representing classes of counterexamples.  SmartCheck has been implemented for Haskell and is freely available.
\end{abstract}

@@

In addition, shrinking is one side of the coin when it comes to making
counterexamples more understandable: the other side is extrapolation from
individual counterexamples to a class of counterexamples.  This leap of
abstraction is often implicitly made by the programmer in determining the reason
why counterexamples fail the property.  For example, the following is a
relatively small counterexample returned when QuickChecking a property in (a
bug-injected version of) Xmonad, a popular X11 window manager written in
Haskell~\cite{xmonad}:
%
\medskip%
\begin{sloppypar}
%\footnotesize
\noindent%
\ttp{StackSet \{current = Screen \{workspace = Workspace \{tag = NonNegative
      \{getNonNegative = 0\}, layout = -1, stack = Just (Stack \{focus = 'S', up =
        "", down = ""\})\}, screen = 1, screenDetail = 1\}, visible = [Screen
    \{workspace = Workspace \{tag = NonNegative \{getNonNegative = 2\}, layout = -1,
        stack = Nothing\}, screen = 2, screenDetail = -1\},Screen \{workspace =
      Workspace \{tag = NonNegative \{getNonNegative = 3\}, layout = -1, stack =
        Nothing\}, screen = 0, screenDetail = -1\}], hidden = [Workspace \{tag =
      NonNegative \{getNonNegative = 1\}, layout = -1, stack = Just (Stack \{focus
        = `\char`\\NUL', up = "", down = ""\})\},Workspace \{tag = NonNegative
      \{getNonNegative = 4\}, layout = -1, stack = Just (Stack \{focus = 'I', up =
        "", down = ""\})\}], floating = fromList []\} }
\end{sloppypar}
\medskip%
\noindent
(This counterexample uses Haskell's default \ttp{Show} instances, which uses
record syntax.)  Programmers may be familiar with having to debug
a ``wall of text'' as shown above.  What if instead a
formula like the following were returned, stating that for any well-typed
values $x_0$, $x_1$, $x_2$, and $x_3$, tested, a counterexample is found?

@@

As shown in the motivating example in Section~\ref{sec:example}, SmartCheck
can shrink more than an order-of-magnitude faster than even the most efficient hand-written
shrink implementations for some properties and programs.  As well, the time
required to shrink scales linearly with the size of the counterexample---other
common shrink implementations often scale in polynomial time or worse.
(Shrinking time is important in real-world uses; see
Section~\ref{sec:experiments} for examples.)

@@

A more clever programmer that understands the semantics of QuickCheck's
\ttp{shrink} implementation\footnote{The following approach was suggested to
  the author by John Hughes.} defines a \ttp{shrink} instance as follows:
%
\begin{code}
shrink (T w i0 i1 i2 i3) = map go xs
  where xs = shrink (w, i0, i1, i2, i3)
        go (w', i0', i1', i2', i3')
           = T w' i0' i1' i2' i3'
\end{code}
%
\noindent
This ``tuple shrink'' definition does not suffer the same shortcomings: it
shrinks a value even if it contains an empty list, and the combinatorial blowup
of shrink candidates is avoided, since a pair \ttp{(a,b)} is shrunk by
attempting to shrink \ttp{a} while holding \ttp{b} constant, then attempting to
shrink \ttp{b} while holding \ttp{a} constant (and is generalized to larger
tuples).  Thus, each argument to \ttp{T} is independently shrunk.

@@

\todo{QC without shrinking Ints}

@@

To motivate the benefits of SmartCheck reduction algorithm, consider
Table~\ref{table:results} and the corresponding graphs in
Figure~\ref{fig:graphs}, giving the results for using QuickCheck and SmartCheck
on the program shown in Figure~\ref{fig:initial} (we omit SmallCheck, since as
noted, it is not even feasible for the example with default instances).  We
graph the final counterexample size and execution time.  Out of 2000 runs, we
show the number of runs resulting in a shrunk counterexample of a given size (by
counting \ttp{N} constructors), and we show the number of runs executing within
a given time limit (in seconds).  Note that we provide two execution-time
graphs, since time scales differ dramatically (we use a logrithmic scale for the
$y$ axis in one graph and a linear scale in the other).  In the table, we show
the mean, median, median, and the results at the 95th percentile values.  (We
give standard deviations, but not that the plots are not necessarily Gaussian.)

@@

%% In regards to generalization, after shrinking the counterexample, SmartCheck
%% will also generalize it.  Note that for any \ttp{Word8} value in the first
%% argument of \ttp{T}, the counterexample still holds.  For example, SmartCheck
%% might return a counterexample of the form
%% %
%% \begin{code}
%% forall x . T x [-20000] [-20000] [] []
%% \end{code}

Finally, these performance results for SmartCheck are fairly generalizable to
other types and properties, since the algorithm is not dependent on the specific
program or property being checked.  However, the results are particularly
pronounced for counterexamples that are particularly ``deep'' or ``broad'' in
terms of their tree structure.  %% For programs and properties for which small
%% counterexamples are easily found, QuickCheck's shrinking is perfectly
%% acceptable.  Furthermore, user-defined shrink methods may vary significantly
%% based on the property and counterexample type, so it is not always clear which
%% implementation to compare against.
We now describe in detail SmartCheck's algorithm for generic shrinking.

@@

class Arbitrary a => SubTypes a where
  size    :: a -> Size
  index   :: a -> Idx -> Maybe SubVal
  replace :: a -> Idx -> SubVal -> a
  constr  :: a -> String
  constrs :: a -> [String]
  opaque  :: a -> Bool
  subVals :: a -> Tree SubVal
\end{code}
%
\noindent
The \ttp{SubTypes} type class requires QuickCheck's \ttp{Arbitrary} as a
super-class.  \ttp{SubTypes} has the following methods:
\begin{itemize}
\item \ttp{size} returns the size of a value---intuitively, the number of
  constructors contained within it,
\item \ttp{index} returns a \emph{sub-value} at a breadth-first index in a
  value,
\item \ttp{replace} replaces a sub-value at a particular focus (returning the
  original value if the index is out-of-bounds),
\item \ttp{constr} returns a string representation of the constructor tagging
  the value,
\item \ttp{constrs} returns the list of all possible constructors from the
  value's type,
\item \ttp{opaque} is false when the type of the value is an ``interesting
  type''; informally, this is a type other than a primitive type like \ttp{Int}
  or \ttp{Char}.  See Section~\ref{sec:base} for a full discussion.
\item \ttp{subVals} returns a tree of all non opaque-type sub-values.  A tree is a
  value with the type
%
  \begin{code}
data Tree a = Node \{ rootLabel :: a
                   , subForest :: [Tree a] \}
  \end{code}
%
\end{itemize}

@@

subVals (B (B L\sub{0} L\sub{1}) L\sub{2}) =
  Node (B L\sub{0} L\sub{1}) [Node L\sub{0} [], Node L\sub{1} []]
       (Node L\sub{1} [])
\end{code}
%
\noindent
The \ttp{SubVal} type is an existential data type, used as a generic container
for sub-values from a counterexample.  We will sometimes refer to the unwrapped
value returned by \ttp{index a i} as the \emph{$i$th sub-value of $a$}, so for
example, (\ttp{B\sub{4} L\sub{6} L\sub{8}}) is the 4th sub-value of \ttp{tree}.
An invariant of \ttp{index} is that for any value \ttp{a}, and for the smallest
$i \geq 0$ such that
%
\begin{code}
index a i == Nothing
\end{code}
%
\noindent
then for all $0 \leq j < i$,
%
\begin{code}
index a j /= Nothing
\end{code}
\noindent
We use this invariant as a termination case in recursive algorithms over the
sub-values of a value.  (Rather than indexes into a data-structure, an
alternative representation is to use a zipper data structure~\cite{zipper} to
traverse data.  We have chosen explicit indexes to write simple tail-recursive
algorithms that can easily be transcribed to imperative languages.)

@@

\section{Shrinking Data}\label{sec:shrinking}
In this section, we describe how to efficiently and generically shrink algebraic
data values.  Recall the basic idea behind the \ttp{shrink}
method of the \ttp{Arbitrary} class: generate a list of values, each of which is
smaller than the current counterexample.  Each of the new values generated
may not bear any relationship to the original counterexample other than being
smaller.  SmartCheck pursues an approach that searches for smaller but
structurally similar counterexamples, as we make precise below.

@@

newVals :: Size -> Int -> SubVal -> IO [SubVal]
newVals sz tries (SubVal a) =
  replicateM tries s where
  s = liftM SubVal (\undef{sizedArbitrary} sz a)

@@

test :: SubTypes a => a -> Idx -> [SubVal]
     -> (a -> Property) -> Maybe a
test cex idx vs prop = go vs
  where
  go []      = Nothing
  go (v:vs') =
    let cex' = replace cex idx v in
    if \undef{pass} prop cex' then go vs'
      else Just cex'
  \end{code}
  \caption{Counterexample reduction algorithm.\label{fig:reduction}}
\end{figure}

Figure~\ref{fig:reduction} shows the reduction algorithm.  In this algorithm and
subsequent algorithms in the paper, functions in {\bf bold font} are left
undefined but their implementation is described in the text.  The function
\ttp{reduce} takes flags to customize the algorithm's behavior, a counterexample
\ttp{cex}, and the property \ttp{prop}.  The reduction begins at the first
proper sub-value of \ttp{cex}; call it \ttp{v}.  When the index \ttp{idx}
becomes out-of-bounds and returns \ttp{Nothing}, the algorithm terminates.
Otherwise, a list of new random values are generated.
%
\begin{code}
sizedArbitrary :: SubTypes a => Size -> a -> IO a
\end{code}
%
\noindent
generates a new value \ttp{v'} having the same type as \ttp{v} and that is
strictly smaller (with respect to the \ttp{size} method) than \ttp{v}.  Just
like QuickCheck's \ttp{arbitrary} method, \ttp{sizedArbitrary} generates
successively larger counterexamples when generating new values with which to
replace a sub-value.

@@

\subsubsection{Variant Counterexample Hypothesis}
A motivation for the design of the reduction algorithm is something we call the
\emph{variant counterexample hypothesis}: in the search space of possible values
from a given type \ttp{T}, if a known counterexample \ttp{cex} is a variant
\ttp{v} of \ttp{T}, then it is most probable that other counterexamples are also
from variant \ttp{v}.  As an example supporting the hypothesis, consider a
property about unintended variable capture over a language's parse tree
represented by a sum type with constructors for module imports, function
definitions, and global-variable assignments, respectively.  A function
definition counterexample can only be reduced to smaller function definition
counterexamples, the only construct in which variable capture is possible.

Recall that the algorithm begins at the \emph{first} sub-value of the
counterexample rather than the 0th sub-value.  The 0th sub-value of \ttp{cex} is
\ttp{cex} itself.  No invariant of the algorithm would be violated by beginning
with the 0th sub-value, and in particular, the algorithm would still terminate.
However, we begin with a strict sub-value because of the hypothesis.

The hypothesis may be incorrect for some properties, in which case SmartCheck
may potentially fail to discover a smaller counterexample.  However, in
Sections~\ref{sec:existential} and~\ref{sec:precondition}, we describe an
approaches to generalize counterexamples based on discovering new counterexample
variants.  These generalization techniques are executed in an (optional)
generalization phase, run after the reduction phase, in which this hypothesis is
implemented.

@@

\todo{reference into with QC, no int shrinking}

@@

  testHole (SubVal a) = do
    a' <- \undef{cast} a :: Maybe a
    if \undef{pass} prop a' then Nothing else Just a'
\end{code}
  \caption{Reduction algorithm with the sub-value counterexample optimization.}
  \label{fig:reduce0}
\end{figure}

@@

\subsection{Universal Sub-Value Generalization}\label{sec:universal}
Consider again the calculator language from Section~\ref{sec:subval}.  The
property \ttp{prop\_div} is violated for any numerator, so we might generalize a
counterexample like
%
\begin{code}
Div (Add (C 7) (C 3)) (Add (C (-5)) (C 5))
\end{code}
%
\noindent
by the formula
%
\begin{code}
forall x . Div x (Add (C (-5)) (C 5))
\end{code}
%
\noindent
since any dividend results in divide-by-zero for the given divisor.  Not only do
the generalizations assist the programmer's insight, but they reduce the sheer
size of the counterexample.  We call the kind of formula just shown
\emph{universal sub-value generalization} and it is implemented in SmartCheck.

@@

\begin{figure}
  \begin{code}
subTrees :: SubTypes a => a -> Idx -> [Idx] -> Bool
subTrees cex idx = any (\undef{subTree} cex idx)

@@

allFail :: SubTypes a => ScArgs -> a -> Idx
  -> [SubVal] -> (a -> Property) -> Bool
allFail args cex idx vs prop =
  length res >= scMinForall args && and res
  where
  res  = mapMaybe go vs
  go   = \undef{fail} prop . replace cex idx
  \end{code}
  \caption{Universal sub-value generation algorithm.}
  \label{fig:universal}
\end{figure}

The extrapolation algorithm is shown in Figure~\ref{fig:universal}; let sketch
its specification.  The algorithm is similar to the reduction algorithm in
Figure~\ref{fig:reduction} (and in the implementation, the algorithms are
generalized and combined).  The function \ttp{extrapolate} returns a list of
indexes to be generalized in the original counterexample.  In the recursive
function \ttp{extrapolate'}, there is a function guard with a call
%
\begin{code}
subTree cex idx0 idx1
\end{code}
%
\noindent
where \ttp{subTree} has the type
%
\begin{code}
subTree :: SubTypes a => a -> Idx -> Idx -> Bool
\end{code}
%
\noindent
The value
%
\begin{code}
subTree cex idx0 idx1
\end{code}
%
is true if in \ttp{cex}, the value at index \ttp{idx0} is a child of index
\ttp{idx1} in a tree representation of \ttp{cex} (i.e., \ttp{subVals cex}).  The
\ttp{subTrees} guard prevents the algorithm from trying to generalize sub-values
that are abstracted away already since their parents have been generalized.
New sub-values are generated by \ttp{newVals}, shown in
Figure~\ref{fig:reduction}.

The function \ttp{allFail} takes a counterexample \ttp{cex}, an index into
\ttp{cex}, a list of new sub-values, and a property.  It returns true if no new
values satisfy the property.  The function
%
\begin{code}
fail :: (a -> Property) -> a -> Maybe Bool
\end{code}
%
\noindent
is roughly the dual of \ttp{pass} in the reduction algorithm: (\ttp{fail
  prop cex}) returns (\ttp{Just True}) if \ttp{cex} passes \ttp{prop}'s
precondition but fails the property; (\ttp{Just False}) if \ttp{cex}
non-trivially satisfies \ttp{prop}, and \ttp{Nothing} if \ttp{cex} fails
\ttp{prop}'s precondition.

Like in the reduction algorithm user-specified flags bound the behavior of the
algorithm.  We bound the size of values to generate by the flag \ttp{scMaxSize},
which is independent of the size of the particular sub-value.  The flag
\ttp{scMaxForall} is the analogue of the \ttp{scMaxReduce} flag, determining the
number of values generated in trying to generalize a value.  The flag
\ttp{scMinForall} is the minimum number of \ttp{Just False} results required
from \ttp{fail} to extrapolate from failed tests to a universal claim.  So, for
example, if \ttp{scMaxForall} is set to 30 and \ttp{scMinForall} is set to 20,
we generate 30 new values, 20 of which must pass the precondition but fail the
property to claim the counterexample can be generalized.

@@

%% Regarding our implementation, we have not profiled the performance of our
%% implementation of SmartCheck, so additional performance gains might be possible.
%% In particular, we perform significant computations over tree-structured data,
%% and a generic zipper~\cite{zipper}, which we do not currently implement, could
%% improve performance.
